---
title: "__TITLE__"
author: "__AUTHOR__"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---

```{r resolve_conflicts, echo=FALSE}
resolve_conflicts <- function(pkg_priority) {
  get_index <- function(pkg_name) {
    idx <- str_which(pkg_priority, pkg_name)

    if(length(idx) == 0) {
      idx <- 0L
    }

    return(idx)
  }

  conflict_lst <- conflict_scout()

  for(func_name in names(conflict_lst)) {
    pkg_index <- map_int(conflict_lst[[func_name]], get_index)

    pkg_index <- pkg_index[pkg_index > 0]

    if(length(pkg_index) == 0) {
      pkg_use <- conflict_lst[[func_name]][1]
    } else {
      pkg_use <- pkg_index %>%
        min() %>%
        pkg_priority[.]
    }

    conflict_prefer(func_name, pkg_use)
  }

  return(conflict_lst)
}
```


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(vctrs)
library(fs)
library(glue)
library(forcats)
library(snakecase)


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )



knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)
```

```{r custom_functions, echo=FALSE}
### Checks if variable is a date/time
is_date <- function(x)
  x %>% inherits(c("POSIXt", "POSIXct", "POSIXlt", "Date", "hms"))


### Returns the category of data type passed to it
categorise_datatype <- function (x) {
  if(all(are_na(x))) return("na")

  if(is_date(x))                          "datetime"
  else if (!is_null(attributes(x)) ||
           all(is_character(x)))          "discrete"
  else if (all(is_logical(x)))            "logical"
  else                                    "continuous"
}


### create_coltype_list() splits columns into various types
create_coltype_list <- function(data_tbl) {
  coltypes  <- data_tbl %>% map_chr(categorise_datatype)
  cat_types <- coltypes %>% unique() %>% sort()

  split_lst <- cat_types %>%
    map(function(x) { coltypes[coltypes %in% x] %>% names() })

  names(split_lst) <- coltypes %>% unique() %>% sort()

  coltype_lst <- list(
    split   = split_lst
   ,columns = coltypes
  )

  return(coltype_lst)
}

```


This workbook was created using the 'dataexpks' template:

https://github.com/DublinLearningGroup/dataexpks



# Introduction

This workbook performs the basic data exploration of the dataset.

```{r set_exploration_params, echo=TRUE}
dataexp_level_exclusion_threshold <- 100

dataexp_cat_level_count <- 40
dataexp_hist_bins_count <- 50
```


# Load Data

First we load the dataset.

```{r load_dataset, echo=TRUE}
### _TEMPLATE_
### Data is loaded into dataset rawdata_tbl here

### We may wish to set column typs
#data_col_types <- cols(
#  VAR1 = col_character(),
#  VAR2 = col_date(),
#  VAR3 = col_number()
#)

### Data is loaded into dataset rawdata_tbl here
#rawdata_tbl <- read_csv(DATAFILE,
#                        locale    = locale(),
#                        col_types = data_col_types,
#                        progress  = FALSE
#                       )

glimpse(rawdata_tbl)
```


## Perform Quick Data Cleaning


```{r perform_simple_datatype_transforms, echo=TRUE}
### _TEMPLATE_
### Do simple datatype transforms and save output in data_tbl
data_tbl <- rawdata_tbl %>% set_colnames(names(.) %>% to_snake_case())

glimpse(data_tbl)
```



```{r, echo=FALSE}
knitr::knit_exit()
```


## Create Derived Variables

We now create derived features useful for modelling. These values are
new variables calculated from existing variables in the data.

```{r create_derived_variables, echo=TRUE}
data_tbl <- data_tbl

glimpse(data_tbl)
```



```{r, echo=FALSE}
knitr::knit_exit()
```


# Perform Basic Checks on Data

We now want to look at some very high level checks on the data, and we leverage
some of the functionality provided by `DataExplorer`.

## Create High-Level Visualisations

We first want to look at a visualisation of some high-level summarys of the
meta-data on this dataset. This gives us a quick view of the categorical and
numeric values in the dataset, as well as the proportions of missing values.

```{r plot_dataexp_introduce}
data_tbl %>%
  plot_intro(
    title   = "High Level Table Summary",
    ggtheme = theme_cowplot()
    )
```


## Check Missing Values

Before we do anything with the data, we first check for missing values
in the dataset. In some cases, missing data is coded by a special
character rather than as a blank, so we first correct for this.

```{r replace_missing_character, echo=TRUE}
### _TEMPLATE_
### ADD CODE TO CORRECT FOR DATA ENCODING HERE
```

With missing data properly encoded, we now visualise the missing data in a
number of different ways.

### Univariate Missing Data

```{r plot_univariate_missing_data, echo=TRUE}
data_tbl %>%
  plot_missing(
    title   = "Summary of Data Missingness",
    group   = list(Good = 0.05, Acceptable = 0.2, Bad = 0.8, Remove = 1),
    ggtheme = theme_cowplot()
    )
```

We now want to repeat this plot but only for those columns that have some
missing values.

```{r plot_univariate_missing_only_data, echo=TRUE}
data_tbl %>%
  plot_missing(
    title        = "Summary of Data Missingness (missing variables only)",
    missing_only = TRUE,
    group        = list(Good = 0.05, Acceptable = 0.2, Bad = 0.8, Remove = 1),
    ggtheme      = theme_cowplot()
    )
```



### Multivariate Missing Data

It is useful to get an idea of what combinations of variables tend to have
variables with missing values simultaneously, so to construct a visualisation
for this we create a count of all the times given combinations of variables
have missing values, producing a heat map for these combination counts.

```{r missing_data_matrix, echo=TRUE}
row_count <- rawdata_tbl %>% nrow()

count_nas <- ~ .x %>% are_na() %>% vec_cast(integer())

missing_plot_tbl <- rawdata_tbl %>%
  mutate(across(everything(), count_nas)) %>%
  mutate(label = pmap_chr(., str_c)) %>%
  group_by(label) %>%
  summarise(across(everything(), sum)) %>%
  arrange(desc(label)) %>%
  select(-label) %>%
  mutate(label_count = pmap_int(., pmax)) %>%
  pivot_longer(
    cols = !label_count,
    names_to  = "col",
    values_to = "count"
    ) %>%
  mutate(
    miss_prop   = count / row_count,
    group_label = sprintf("%6.4f", round(label_count / row_count, 4))
    )

ggplot(missing_plot_tbl %>% slice_max(order_by = miss_prop, n = 10)) +
  geom_tile(aes(x = col, y = group_label, fill = miss_prop), height = 0.8) +
  scale_fill_continuous() +
  scale_x_discrete(position = "top") +
  xlab("Variable") +
  ylab("Missing Value Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

This visualisation takes a little explaining.

Each row represents a combination of variables with simultaneous missing
values. For each row in the graphic, the coloured entries show which particular
variables are missing in that combination. The proportion of rows with that
combination is displayed in both the label for the row and the colouring for
the cells in the row.


## Inspect High-level-count Categorical Variables

With the raw data loaded up we now remove obvious unique or near-unique
variables that are not amenable to basic exploration and plotting.

```{r find_highlevelcount_categorical_variables, echo=TRUE}
coltype_lst <- create_coltype_list(data_tbl)

count_levels <- ~ .x %>% unique() %>% length()

catvar_valuecount_tbl <- data_tbl %>%
  summarise(across(coltype_lst$split$discrete, count_levels)) %>%
  pivot_longer(
    cols      = everything(),
    names_to  = "var_name",
    values_to = "level_count"
    ) %>%
  arrange(desc(level_count))

print(catvar_valuecount_tbl)

row_count <- data_tbl %>% nrow()

cat(glue("Dataset has {row_count} rows\n"))
```

Now that we a table of the counts of all the categorical variables we can
automatically exclude unique variables from the exploration, as the level
count will match the row count.

```{r remove_id_variables, echo=TRUE}
unique_vars <- catvar_valuecount_tbl %>%
  filter(level_count == row_count) %>%
  pull(var_name)

print(unique_vars)

explore_data_tbl <- data_tbl %>%
  select(-one_of(unique_vars))
```

Having removed the unique identifier variables from the dataset, we
may also wish to exclude categoricals with high level counts also, so
we create a vector of those variable names.

```{r collect_highcount_variables, echo=TRUE}
highcount_vars <- catvar_valuecount_tbl %>%
  filter(level_count >= level_exclusion_threshold,
         level_count < row_count) %>%
  pull(var_name)

cat(str_c(highcount_vars, collapse = ", "))
```

We now can continue doing some basic exploration of the data. We may
also choose to remove some extra columns from the dataset.

```{r drop_variables, echo=TRUE}
### You may want to comment out these next few lines to customise which
### categoricals are kept in the exploration.
drop_vars <- c(highcount_vars)

if (length(drop_vars) > 0) {
  explore_data_tbl <- explore_data_tbl %>%
      select(-one_of(drop_vars))

  cat(str_c(drop_vars, collapse = ", "))
}
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


# Univariate Data Exploration

Now that we have loaded the data we can prepare it for some basic data
exploration.


## Quick Univariate Data Summaries

We use a number of summary visualisations provided by `DataExplorer`: a
facet plot across each variable with categorical variables getting bar plots
and numerical plots getting histograms.

We first look at the barplots of categorical variables.

```{r plot_dataexp_bar, message=TRUE, echo=TRUE}
data_tbl %>%
  plot_bar(
    ncol    = 4,
    nrow    = 3,
    title   = "Barplots of Data",
    ggtheme = theme_cowplot()
    )
```


We then have a quick look at histograms of the numeric variables.

```{r plot_dataexp_hist, message=TRUE, echo=TRUE}
data_tbl %>%
  plot_histogram(
    ncol    = 4,
    nrow    = 3,
    title   = "Histograms of Data",
    ggtheme = theme_cowplot()
    )
```


Finally, we split the remaining variables into different categories and then
produce a sequence of plots for each variable.


```{r separate_exploration_cols, echo=TRUE}
coltype_lst <- create_coltype_list(explore_data_tbl)

print(coltype_lst)
```


## Logical Variables

Logical variables only take two values: TRUE or FALSE. It is useful to see
missing data as well though, so we also plot the count of those.

```{r create_univariate_logical_plots, echo=TRUE, warning=FALSE}
logical_vars <- coltype_lst$split$logical %>% sort()

for (plot_varname in logical_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  na_count <- explore_data_tbl %>% pull(.data[[plot_varname]]) %>% are_na() %>% sum()

  plot_title <- glue("Barplot of Counts for Variable: {plot_varname} ({na_count} missing values)")

  explore_plot <- ggplot(explore_data_tbl) +
    geom_bar(aes(x = .data[[plot_varname]])) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  plot(explore_plot)
}
```


## Numeric Variables

Numeric variables are usually continuous in nature, though we also have
integer data.

```{r create_univariate_numeric_plots, echo=TRUE, warning=FALSE}
numeric_vars <- coltype_lst$split$continuous %>% sort()

for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  plot_var <- explore_data_tbl %>% pull(.data[[plot_varname]])
  na_count <- plot_var %>% are_na() %>% sum()

  plot_var %>% summary() %>% print()

  plot_title <- glue("Histogram Plot for Variable: {plot_varname} ({na_count} missing values)")

  neg_data_tbl <- explore_data_tbl %>%
    filter(.data[[plot_varname]] < 0) %>%
    mutate(var_val = abs(.data[[plot_varname]]))

  pos_data_tbl <- explore_data_tbl %>%
    filter(.data[[plot_varname]] >= 0) %>%
    mutate(var_val = abs(.data[[plot_varname]]))

  all_plot <- ggplot(explore_data_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = hist_bins_count) +
    geom_vline(xintercept = mean(plot_var, na.rm = TRUE),
               colour = "red",   size = 1.5) +
    geom_vline(xintercept = median(plot_var, na.rm = TRUE),
               colour = "green", size = 1.5) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(
      plot_title,
      subtitle = "(red line is mean, green line is median)"
      )


  pos_log_plot <- ggplot(pos_data_tbl) +
    geom_histogram(aes(x = var_val), bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Positive Values")

  neg_log_plot <- ggplot(neg_data_tbl) +
    geom_histogram(aes(x = var_val), bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Negative Values")


  plot_grid(
      all_plot,
      NULL,
      pos_log_plot,
      neg_log_plot,
      nrow = 2
      ) %>%
    print()
}
```

## Categorical Variables

Categorical variables only have values from a limited, and usually fixed,
number of possible values

```{r create_univariate_categorical_plots, echo=TRUE, warning=FALSE}
categorical_vars <- coltype_lst$split$discrete %>% sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  na_count <- explore_data_tbl %>% pull(.data[[plot_varname]]) %>% are_na() %>% sum()

  plot_title <- glue("Barplot of Counts for Variable: {plot_varname} ({na_count} missing values)")

  standard_plot_tbl <- explore_data_tbl %>%
    count(.data[[plot_varname]]) %>%
    mutate({{plot_varname}} := fct_relabel(.data[[plot_varname]], str_trunc, width = 15))

  standard_plot <- ggplot(standard_plot_tbl) +
    geom_bar(aes(x = .data[[plot_varname]], weight = n)) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  standard_plot %>% print()


  desc_plot_tbl <- explore_data_tbl %>%
    pull(.data[[plot_varname]]) %>%
    fct_lump(n = cat_level_count) %>%
    fct_count() %>%
    mutate(f = fct_relabel(f, str_trunc, width = 15))

  desc_plot <- ggplot(desc_plot_tbl) +
    geom_bar(aes(x = fct_reorder(f, -n), weight = n)) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  desc_plot %>% print()
}
```


## Date/Time Variables

Date/Time variables represent calendar or time-based data should as time of the
day, a date, or a timestamp.

```{r create_univariate_datetime_plots, echo=TRUE, warning=FALSE}
datetime_vars <- coltype_lst$split$datetime %>% sort()

for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  plot_var <- explore_data_tbl %>% pull(.data[[plot_varname]])
  na_count <- plot_var %>% are_na() %>% sum()

  plot_var %>% summary() %>% print()

  plot_title <- glue("Barplot of Dates/Times in Variable: {plot_varname} ({na_count} missing values)")


  explore_plot <- ggplot(explore_data_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title)

  plot(explore_plot)
}
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


# Bivariate Data Exploration

We now move on to looking at bivariate plots of the data set.

## Facet Plots on Variables

A natural way to explore relationships in data is to create univariate
visualisations facetted by a categorical value.

```{r bivariate_facet_data, echo=TRUE}
### _TEMPLATE_
### facet_varname <- ''
facet_varname <- coltype_lst$split$discrete[1]

dataexp_facet_count_max <- 3
```


### Logical Variables

For logical variables we facet on barplots of the levels, comparing TRUE,
FALSE and missing data.

```{r create_bivariate_logical_plots, echo=TRUE}
logical_vars <- logical_vars[!logical_vars %in% facet_varname] %>% sort()


for (plot_varname in logical_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl %>% filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = .data[[plot_varname]])) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  plot(explore_plot)
}
```


### Numeric Variables

For numeric variables, we facet on histograms of the data.

```{r create_bivariate_numeric_plots, echo=TRUE}
for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl %>% filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_continuous(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  print(explore_plot)
}
```

### Categorical Variables

We treat categorical variables like logical variables, faceting the barplots
of the different levels of the data.

```{r create_bivariate_categorical_plots, echo=TRUE}
categorical_vars <- categorical_vars[!categorical_vars %in% facet_varname] %>% sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl %>%
    filter(!are_na(.data[[plot_varname]])) %>%
    mutate(
      varname_trunc = fct_relabel(.data[[plot_varname]], str_trunc, width = 10)
      )

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = varname_trunc)) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

  plot(explore_plot)
}
```


### Date/Time Variables

Like the univariate plots, we facet on histograms of the years in the dates.

```{r create_bivariate_datetime_plots, echo=TRUE}
for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl %>% filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}"))

  plot(explore_plot)
}
```

```{r free_memory_facetplot, echo=FALSE}
rm(plot_var, plot_tbl)
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


# Custom Explorations

In this section we perform various data explorations.


# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
